Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-02-13 response_letter.txt

Overall, I recommend to accept the paper. The topic and the treatment
with dependency networks make it a timely contribution. The paper is
reasonably well-written.

Specific suggestions for improvements:

-Page 1: "The overall complexity ... grows like Theta(4^gamma) where gamma
is a parameter whose value is bounded only by the length" This seems
rather strange because if gamma is the overall length, then 4^gamma can be
trivially obtained by enumerating all sequences and then computing the
energies for each sequence. Maybe it is supposed to mean that in the
worst-case gamma can be the length of the sequence? But this is also true
for the treewidth making this remark maybe a bit unfair. I suggest to be
more precise and explain roughly what gamma is.

>>>>> TODO: Elaborate that gamma=n is worst case and in practice much
      smaller due to ear decomp (similar to tree width)

@@

-Page 2 "in a principled way" Could you provide some further explanation
what "principled" means here?

>>>>> TODO --> YANN; something like 'generic'?

-Page 3 The definition of RNA structure allows that one position i takes
part in more than one base pair (e.g. {(1,2),(1,3)} is a set of base pairs
and thus a secondary structure according to the definition).

>>>>> TODO add this restriction (we do not need degree>1, even if it does not harm)

@@

-Page 3 "(Si,Sj)" -> "{Si,Sj}"

-Page 3 Please highlight a bit better when the compatability graph is of
interest and when the dependency graph is of interest. It seems that the
compatability graph is only interesting in Section 3?

-Page 3 It seems strange that the partition function depends on S while S
is used in the sum. Should it be "Z_E_R(n)" or "Z_E_R(S_n)"?

>>>> DONE. Embarassing typo

-Page 3 "as the minimal set" Is it necessarily unique? If not, "as a
minimal set".
DONE

-Page 3 "an hypergraph" -> "a hypergraph"
DONE

-Page 4 "base pair models" -> "base pair model"
DONE

-Page 4 "ins polynomial time" -> "in polynomial time"
DONE

-Page 4 "strongly connected" Why "strongly" is the graph directed? There
is no discussion of weak vs. strong connected components.
DONE

@@

-Page 4,5 two missing "\qed"s

-Page 5 In the definition of "f[a]" one needs that "dep(f)" contains "dom(A)"
DONE

-Page 5 The difference variables should probably be defined only for
adjacent bags of the tree decomposition? Otherwise, I don't see how the
claim about the one-variable difference is true.
DONE

-Page 5 "Denote by A(Y)" What is Y?
DONE

-Page 6 In Line 6 of the algorithm, I don't see how m_w->u has an entry
for "[a|a']" as "a" may contain some new variables.
TODO

-Page 6 "Its correctness" -> "The correctness of the algorithm"
DONE

@@

IGNORE this for now.

-Page 6 "requires O(..." + "time"
DONE

-Page 7 For Theorem 1, I was wondering whether it is really necessary to
consider the total number of connected components. It seems that only the
maximum number of components per bag is crucial?

TODO: obviously, we confused the reviewer---
MOREOVER: strategy should be optimized. Instead of fixing all polarities in the beginning,
it suffices to fix the polarities simultaneously in one bag....

-Page 7 "we encode the network energy ..." I was a bit confused here. What
is w_l? Is it possible to give the formula?
>>>>>> CHECK

-Page 8 Should it "ln" instead of "log"?
DONE

-Page 8 What is "gamma" in "gamma^{mu_l}"?
DONE

-Page 8 "expected number grows like Theta(1)" This sounds a bit awkward as
Theta(1) doesn't really grow. How about "expected number is Theta(1)"?
DONE

-Page 9 In the definition of f(S), the first sum should be from l=1?
DONE

-Page 14 Lemma 1 is a well-known fact about tree decompositions. It
does not need a proof and should be found in some surveys about
treewidth.  >>>> -> YANN: add that this is a folklore results, which
we show for the sake of completeness

### 2018-02-13 RNARedPrint-Bioinfo2018.tex
Edits due to comments of reviewer 2

@@

\newcommand{\TODO}[1]{\Final{\begin{mdframed}[leftline=false,rightline=false,linecolor=red!70!black,linewidth=1.5pt]\color{red!70!black}\sf TODO: #1\end{mdframed}}}
\newcommand{\parHead}[1]{\Final{\paragraph{#1}}}

@@

\TODO{Develop questions (relevance, difficulty...). Maybe tease results?}
\parHead{Motivation.} The exponential time and space requirements of the \Software{RNAdesign} method already raise the question of the \Def{complexity of (uniform) sampling for multi-target design}. Since stochastic backtrack can be performed in linear time per sample, the method is dominated by the precomputation step, which requires counting valid designs. Thus, we focus on the question: \emph{Is there a polynomial-time algorithm to count valid multi-target designs?} In Section~\ref{sec:counting}, we answer in the negative, showing that there exists no such algorithm unless ${\sf P}={\sf NP}$. Our result relies on a surprising bijection (up to a trivial symmetry) between valid sequences  and independent sets of a bipartite graph, being the object of recent breakthroughs in approximate counting complexity~\cite{Bulatov2013,Cai2016}.
The hardness of counting (and conjectured hardness of sampling) does not preclude, however, practically applicable algorithms for counting and sampling. In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, in a principled way, from a Boltzmann distribution based on expressive energy models? How to enforce additional constraints, such as the \GCb-content, complex sequence constraints, or the energy of individual structures?}

@@

%To provide a simple concrete example of multi-target design, consider designing a single sequence where all target structures are as stable as possible.
%That is, one asks for the sequence $S$ that minimizes the energy differences of the target structures to the minimum energy for this sequence among all possible structures; one aims to minimize
%   \begin{equation}
%\label{eq:design-objective}
%\sum_{\ell\in[1,k]} E(S;R_\ell) - E_0(S),
%\end{equation}
%over all sequences $S$, where $E_0(S):=\min_R E(S;R)$.
%\TODO{even Taneda's objective is significantly more complex; how much do we want to talk about this...}
%
%The proposed Boltzmann-weighted sampling approach can support this design procedure by sampling start structures from the Boltzmann distribution over
%\begin{equation}
%\label{eq:sampling-objective} \sum_{\ell\in[1,k]} w_\ell E(S;R_\ell)
%\end{equation}
%or extensions thereof, e.g.~for controlling base frequencies. Suitable weights $w_\ell$ to accommodate the design objective can be approximated automatically after the machinery for Boltzmann sampling over such functions is set up.
%
%\TODO{Rewrite (from here to "Contributions"): sounds too negative}
%Notably, by Eq.~(\ref{eq:sampling-objective}) we do not suggest to sample over the energy difference of Eq.~$(\ref{eq:design-objective})$. The term $\min_R E(S;R)$, which---for each sequence---minimizes over the entire structure space, precludes efficient sampling even by the proposed approach.
%
%Already sampling over Eq.~(\ref{eq:sampling-objective}) strictly extends the only previously existing approach to controlled sampling from single to multi-target sampling (in complete analogy to IncaRNAtion only after GC content is controlled, which we present later) and subsumes uniform sampling for multi-target design.
%
%Naturally, also more complex objective functions are desirable for RNA design and have been discussed in the literature; e.g. the ensemble defect~\cite{Zadeh:2010}. Typically, more complex objective functions make design computationally much more challenging. Similarily, also the proposed sampling framework supports only a limited class of objective functions directly and efficiently---namely ``only'' all additive energy functions. Nevertheless, one can safely assume that the design under various more realistic objective functions profits from controlled sampling as long as the sampling is directed towards the final objective function.

@@

%We solve the problem efficiently for any fixed parameter $w$, whose value reflects the intrinsic complexity of the structures and energy function $E_\R$.
%
As will be explicited in subsequent sections, our approach relies on breaking down the energy function $E_\R(S)$ into additive components, each depending on only few sequence positions. Given $\R$, we express $E_\R(S)$
as the sum of energy contributions $f(S)$ over a set $\F$ (of
functions $f:\S_n\to\real$), s.t.~$E_\R(S)=\sum_{f\in\F} f(S)$. This
captures realistic RNA energy models, while bounding the
dependencies to sequence positions introduced by each single $f$. Formally, define the \Def{dependencies} $\dep(f)$ of $f$ as the minimal set of sequence positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all sequences $S$ and $S'$ agreeing at all positions in $\I$.

@@

%%% reformulate later
%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$. Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair models,with $\beta\neq 1$. As previously noted~\cite{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides. The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.

@@

To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are  strongly connected and bipartite ($(V_1\times V_2) \cap E = \varnothing$), noting that hardness over restricted instances implies hardness of the general problem. Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$. For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ (resp. $V_2$) to only $\Ab$ and $\Gb$ (resp. $\Cb$ and $\Ub$). Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.

@@

A \Def{(partial) assignment} is a function $\val: \X' \to \Sigma$
that maps the variables $\X'\subseteq \X$ to bases; we call $\X'$ the \Def{domain of the assignment
  $\val$}, written $\dom(\val)$; $\val$ is \Def{total} iff $\dom(\val)=\X.$
%
For assignments $\val$ and functions $f$, we define the
special notation $\evalfor{f}{\val}$ to \Def{evaluate $f$ for $\val$}; i.e.{}
$\evalfor{f}{\val} := f(S),$ for any sequence $S$ where $S_i=\val(i)$ for all
$i\in\dep(f).$ Then, the \Def{energy $\energy{\network}(S)$ of a
  sequence} $S$ in a network $\network$ is defined as sum of the
values of all functions in $f\in\F$ evaluated for $\val_S$, i.e.
$\energy{\network}(S) := \sum_{f\in F} \evalfor{f}{\val_S}.$

@@

Let us now introduce some definitions. A \Def{cluster tree} for the network $\network=(\X,\F)$ is a tuple $(T,\chi,\phi)$, where $(T,\chi)$ is a tree decomposition of $G_\F$, and $\phi(v)$ represents a set of functions $f$, each uniquely assigned to a node $v\in T$; $\vars(f)\subseteq\chi(v)$ and $\phi(v)\cap \phi(v')=\varnothing$ for all $v\neq v'$.
For two nodes $v$ and $u$ of the cluster tree, define their \Def{separator} as $\separator{u}{v} := \chi(u)\cap\chi(v)$; moreover, we define the \Def{difference variables} from $u$ to $v$ by $\difference{u}{v}:=\chi(v) - \separator{u}{v}$.

Denote by $\assignments(\Y)$ the set of all assignments of the variables in $\Y$; moreover, given non-overlapping assignments $\val'$ and $\val''$, we write $\substitute{\val'}{\val''}$ to denote the assignment defined as $\val'$ and $\val''$ respectively on $\dom(\val')$ and $\dom(\val'')$. 
Finally we assume, w.l.o.g., that all variable difference sets $\difference{u}{v}$ are restricted to a single variable: for any given cluster tree, an equivalent (in term of treewidth) cluster tree can always be obtained by inserting at most $\Theta(|\X|)$ additional clusters.

Let us now consider the \Def{computation of the partition function}.
Given is the RNA design network $\network=(\X,\F)$ and its cluster
tree decomposition $(T,\chi,\phi)$.  W.l.o.g, we assume that $T$ is
connected and contains a dedicated node $r$, with $\chi(r)=\varnothing$
and $\phi(r)=\varnothing$, added as a virtual root
$r$ connected to a node in each connected
component of $T$.  Now, we consider the set of directed edges
$\edgesToR{}$ of $T$ oriented to $r$; define $T_r(u)$ as the induced
subtree of $u$. Algorithm~\ref{alg:pf} computes the partition function by
passing messages along these directed edges $u\to v$ (i.e. always from
some child $u$ to its parent $v$). Each message is a function on
variables $\vars(m)\subseteq \X$ to $\real\cup\{\infty\}$. The message
from $u$ to $v$ represents the partition functions of the subtree of
$u$ for all possible assignments to the variables
$\vars(m)=\separator{u}{v}$. Its correctness can be shown by induction over $T$ (Appendix~\ref{appsec:correctness}).
After running Alg.~\ref{alg:pf}, 
multiplying the 0-ary messages sent to the root $r$ yields the total partition function:
\begin{math}
  \partfun{\network} = \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

For our analysis, we define the \emph{maximum separator size} $s$ as
$\max_{u,v\in V} | \separator{u}{v} |$ and denote the maximum size of
$\difference{u}{v}$ over $(u,v)\in\edgesToR{}$ as $D$.  Generally,
without specific optimizations, running Alg.~\ref{alg:pf} requires
$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ and $\mathcal{O}(|V|\cdot4^s)$ space (Appendix~\ref{appsec:algcomplexity});
Alg.~\ref{alg:sampling} would require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per
sample on arbitrary tree decompositions (Appendix~\ref{appsec:algcomplexity}). W.l.o.g. we assume that
$D=1$; note that tree decompositions can generally be transformed,
such that $\difference{u}{v}\leq 1$.
%
Moreover, the size of $\F$ is linearly bounded: for $k$
input structures for sequences of length $n$, the energy function is
expressed by $\mathcal{O}(n\,k)$ functions. Finally, the number of cluster
tree nodes is in $O(n)$, such that $|\F|+|V| \in \mathcal{O}(n\,k)$.

@@

Multidimensional Boltzmann sampling requires the ability to \Def{sample from a weighted distribution} over the set of compatible sequences, where the probability of a sequence $S$ with free energies $(E_1,\ldots,E_k)$ for its target structures is
$\mathbb{P}(S\mid \pmb{\pi}) = \frac{\prod_{\ell=1}^{k} \pi_i^{-E_i}}{\partfun{\pmb{\pi}}},$
where $\pmb{\pi}:=(\pi_1\cdots\pi_k)$ is a vector of positive real-valued \Def{weights}, and $\partfun{\pmb{\pi}}$ is the weighted partition function. Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \log(\pi_\ell)\, E(\val)/\beta$. The probability of a sequence $S$ is thus proportional to 
$ \prod_{\ell=1}^{k} e^{-\log(\pi_\ell)\, E_i} = \prod_{\ell=1}^{k} \pi_i^{-E_i}. $

One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, \emph{i.e.} such that  $\mathbb{E}(E_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$.
The expected value of $E_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see App.~\ref{sec:weight_derivatives}). More generally, computing a suitable $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\cite{Denise2010,Bendkowski2017}. 
In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(exp(\beta),\dots,\exp(\beta))$ and, at each iteration, generates a sample $\mathcal{S}$ of sequences. The expected value of an energy $E_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}E_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$. 
$\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$.
While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior.

A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}. The rationale for a rejection approach follows from: 
i)\emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing Cupal \emph{et al}~\cite{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}). 
%Indeed, as shown in Figure~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
For base pair-based energy models and special compatibility graph (paths, cycles\ldots) this property rigorously follows from analytic combinatorics, see Bender-Richmond-Williamson~\cite{Bender1983} and Drmota~\cite{Drmota1997}. In such cases, the expected number of rejections before reaching the targeted energies grows like $\Theta(1)$ when $\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when $\varepsilon=0$. The \GCb-content of designs can also be controlled, jointly and in a similar way, as done in \Software{IncaRNAtion}~\cite{Reinharz2013}.

@@

For a multi-target design instance, individual reasonable target energies are estimated from averaging the energy over samples at relatively high weights $e^{\beta}$, controlling the \GCb-content at 50\%. For each instance of the \Software{Modena} benchmark~\cite{Taneda2015}, we estimated such target energies and subsequently generated $1\,000$ seed sequences at these energies.
As its rationale, this procedure is tailored to produce sequences with similar Turner energy that favor the stability of the target structures having moderate \GCb-content; all these properties are desirable objectives for RNA design.
All sampled sequences are evaluated under the multi-stable design objective function of \cite{Hammer2017}:
\vspace{-12pt}
\begin{align}
  \label{eq:blueprintobjective}
  f(S)= \quad & \frac{1}{k} \sum_\ell^{k} (E(S, R_\ell) - G(S))\notag\\
   & +\ \xi \frac{2}{k(k-1)} \sum\limits_{\ell<j}^k|E(S,R_\ell) - E(S,R_j)|.
\end{align}

>>>>

\usepackage{xcolor}

@@

\newcommand{\TODO}[1]{\textcolor{red!70!black}{\textbf{TODO: #1}}}

\newcommand{\parHead}[1]{\Final{\paragraph{#1}}}

@@

\parHead{Motivation.} The exponential time and space requirements of the \Software{RNAdesign} method already raise the question of the \Def{complexity of (uniform) sampling for multi-target design}. Since stochastic backtrack can be performed in linear time per sample, the method is dominated by the precomputation step, which requires counting valid designs. Thus, we focus on the question: \emph{Is there a polynomial-time algorithm to count valid multi-target designs?} In Section~\ref{sec:counting}, we answer in the negative, showing that there exists no such algorithm unless ${\sf P}={\sf NP}$. Our result relies on a surprising bijection (up to a trivial symmetry) between valid sequences  and independent sets of a bipartite graph, being the object of recent breakthroughs in approximate counting complexity~\cite{Bulatov2013,Cai2016}.
The hardness of counting (and conjectured hardness of sampling) does not preclude, however, practically applicable algorithms for counting and sampling. In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, in a principled way, from a Boltzmann distribution based on expressive energy models? How to enforce additional constraints, such as the \GCb-content, complex sequence constraints, or the energy of individual structures?}

@@

%To provide a simple concrete example of multi-target design, consider designing a single sequence where all target structures are as stable as possible.
%That is, one asks for the sequence $S$ that minimizes the energy differences of the target structures to the minimum energy for this sequence among all possible structures; one aims to minimize
%   \begin{equation}
%\label{eq:design-objective}
%\sum_{\ell\in[1,k]} E(S;R_\ell) - E_0(S),
%\end{equation}
%over all sequences $S$, where $E_0(S):=\min_R E(S;R)$.
%
%The proposed Boltzmann-weighted sampling approach can support this design procedure by sampling start structures from the Boltzmann distribution over
%\begin{equation}
%\label{eq:sampling-objective} \sum_{\ell\in[1,k]} w_\ell E(S;R_\ell)
%\end{equation}
%or extensions thereof, e.g.~for controlling base frequencies. Suitable weights $w_\ell$ to accommodate the design objective can be approximated automatically after the machinery for Boltzmann sampling over such functions is set up.
%
%\TODO{Rewrite (from here to "Contributions"): sounds too negative}
%Notably, by Eq.~(\ref{eq:sampling-objective}) we do not suggest to sample over the energy difference of Eq.~$(\ref{eq:design-objective})$. The term $\min_R E(S;R)$, which---for each sequence---minimizes over the entire structure space, precludes efficient sampling even by the proposed approach.
%
%Already sampling over Eq.~(\ref{eq:sampling-objective}) strictly extends the only previously existing approach to controlled sampling from single to multi-target sampling (in complete analogy to IncaRNAtion only after GC content is controlled, which we present later) and subsumes uniform sampling for multi-target design.
%
%Naturally, also more complex objective functions are desirable for RNA design and have been discussed in the literature; e.g. the ensemble defect~\cite{Zadeh:2010}. Typically, more complex objective functions make design computationally much more challenging. Similarily, also the proposed sampling framework supports only a limited class of objective functions directly and efficiently---namely ``only'' all additive energy functions. Nevertheless, one can safely assume that the design under various more realistic objective functions profits from controlled sampling as long as the sampling is directed towards the final objective function.

@@

%We solve the problem efficiently for any fixed parameter $w$, whose value reflects the intrinsic complexity of the structures and energy function $E_\R$.
%
As will be explicited in subsequent sections, our approach relies on breaking down the energy function $E_\R(S)$ into additive components, each depending on only few sequence positions. Given $\R$, we express $E_\R(S)$
as the sum of energy contributions $f(S)$ over a set $\F$ (of
functions $f:\S_n\to\real$), s.t.~$E_\R(S)=\sum_{f\in\F} f(S)$. This
captures realistic RNA energy models, while bounding the
dependencies to sequence positions introduced by each single $f$. Formally, define the \Def{dependencies} $\dep(f)$ of $f$ as a minimal set of sequence positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all sequences $S$ and $S'$ agreeing at all positions in $\I$.

@@

%%% reformulate later
%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$. Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$. As previously noted~\cite{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides. The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.

@@

To establish that claim, we consider instances $G=(V_1\cup V_2, E)$ that are connected and bipartite ($(V_1\times V_2) \cap E = \varnothing$), noting that hardness over restricted instances implies hardness of the general problem. Moreover remark that, as observed in Subsec.~\ref{sec:complexity}, assigning a nucleotide to a position $u\in V$ constrains the parity ($\{\Ab,\Gb\}$ or $\{\Cb,\Ub\}$) of all positions in the connected component of $u$. For this reason, we restrict our attention to the counting of valid designs \emph{up to trivial} $(\Ab\leftrightarrow \Cb/\Gb\leftrightarrow \Ub)$ symmetry, by constraining the positions in $V_{1}$ (resp. $V_2$) to only $\Ab$ and $\Gb$ (resp. $\Cb$ and $\Ub$). Let $\Design{G}$ denote the subset of all designs for $G$ under this constraint, noting that $\NumDesign(G) = 2\cdot|\Design{G}|$.

@@

A \Def{(partial) assignment} is a function $\val: \X' \to \Sigma$
that maps the variables $\X'\subseteq \X$ to bases; we call $\X'$ the \Def{domain of the assignment
  $\val$}, written $\dom(\val)$; $\val$ is \Def{total} iff $\dom(\val)=\X.$
%
For assignments $\val$ and functions $f$, $\dep(f)\subseteq\dom(\val)$,
we define the special notation $\evalfor{f}{\val}$ to \Def{evaluate
  $f$ for $\val$}; i.e.{} $\evalfor{f}{\val} := f(S),$ for any
sequence $S$ where $S_i=\val(i)$ for all $i\in\dep(f).$ Then, the
\Def{energy $\energy{\network}(S)$ of a sequence} $S$ in a network
$\network$ is defined as sum of the values of all functions in
$f\in\F$ evaluated for $\val_S$, i.e.
$\energy{\network}(S) := \sum_{f\in F} \evalfor{f}{\val_S}.$

@@

Let us now introduce some definitions. A \Def{cluster tree} for the
network $\network=(\X,\F)$ is a tuple $(T,\chi,\phi)$, where
$(T,\chi)$ is a tree decomposition of $G_\F$, and $\phi(v)$ represents
a set of functions $f$, each uniquely assigned to a node $v\in T$;
$\vars(f)\subseteq\chi(v)$ and $\phi(v)\cap \phi(v')=\varnothing$ for
all $v\neq v'$.  For two nodes $v$ and $u$ of the cluster tree, define
their \Def{separator} as $\separator{u}{v} := \chi(u)\cap\chi(v)$;
moreover, we define the \Def{difference variables} from $u$ to an
adjacent $v$ by $\difference{u}{v}:=\chi(v) - \separator{u}{v}$.

For a set $Y$ of variables, denote the set of all assignments of the
variables in $\Y$ by $\assignments(\Y)$; moreover, given
non-overlapping assignments $\val'$ and $\val''$, we write
$\substitute{\val'}{\val''}$ to denote the assignment defined as
$\val'$ and $\val''$ respectively on $\dom(\val')$ and $\dom(\val'')$.
Finally we assume, w.l.o.g., that all variable difference sets
$\difference{u}{v}$ are restricted to a single variable: for any given
cluster tree, an equivalent (in term of treewidth) cluster tree can
always be obtained by inserting at most $\Theta(|\X|)$ additional
clusters.

Let us now consider the \Def{computation of the partition function}.
Given is the RNA design network $\network=(\X,\F)$ and its cluster
tree decomposition $(T,\chi,\phi)$.  W.l.o.g, we assume that $T$ is
connected and contains a dedicated node $r$, with $\chi(r)=\varnothing$
and $\phi(r)=\varnothing$, added as a virtual root
$r$ connected to a node in each connected
component of $T$.  Now, we consider the set of directed edges
$\edgesToR{}$ of $T$ oriented to $r$; define $T_r(u)$ as the induced
subtree of $u$. Algorithm~\ref{alg:pf} computes the partition function by
passing messages along these directed edges $u\to v$ (i.e. always from
some child $u$ to its parent $v$). Each message is a function on
variables $\vars(m)\subseteq \X$ to $\real\cup\{\infty\}$. The message
from $u$ to $v$ represents the partition functions of the subtree of
$u$ for all possible assignments to the variables
$\vars(m)=\separator{u}{v}$. The correctness of the algorithm can be shown by induction over $T$ (Appendix~\ref{appsec:correctness}).
After running Alg.~\ref{alg:pf}, 
multiplying the 0-ary messages sent to the root $r$ yields the total partition function:
\begin{math}
  \partfun{\network} = \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

For our analysis, we define the \emph{maximum separator size} $s$ as
$\max_{u,v\in V} | \separator{u}{v} |$ and denote the maximum size of
$\difference{u}{v}$ over $(u,v)\in\edgesToR{}$ as $D$.  Generally,
without specific optimizations, running Alg.~\ref{alg:pf} requires
$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and $\mathcal{O}(|V|\cdot4^s)$ space (Appendix~\ref{appsec:algcomplexity});
Alg.~\ref{alg:sampling} would require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per
sample on arbitrary tree decompositions (Appendix~\ref{appsec:algcomplexity}). W.l.o.g. we assume that
$D=1$; note that tree decompositions can generally be transformed,
such that $\difference{u}{v}\leq 1$.
%
Moreover, the size of $\F$ is linearly bounded: for $k$
input structures for sequences of length $n$, the energy function is
expressed by $\mathcal{O}(n\,k)$ functions. Finally, the number of cluster
tree nodes is in $O(n)$, such that $|\F|+|V| \in \mathcal{O}(n\,k)$.

@@

Multidimensional Boltzmann sampling requires the ability to \Def{sample from a weighted distribution} over the set of compatible sequences, where the probability of a sequence $S$ with free energies $(E_1,\ldots,E_k)$ for its target structures is
$\mathbb{P}(S\mid \pmb{\pi}) = \frac{\prod_{\ell=1}^{k} \pi_i^{-E_i}}{\partfun{\pmb{\pi}}},$
where $\pmb{\pi}:=(\pi_1\cdots\pi_k)$ is a vector of positive real-valued \Def{weights}, and $\partfun{\pmb{\pi}}$ is the weighted partition function. Such a distribution can be induced by a simple modification of the functions described in Sec.~\ref{sec:energy_models}, where any energy function $E(\val)$ for a structure $\ell$ is replaced by $E'(\val):= \ln(\pi_\ell)\, E(\val)/\beta$. The probability of a sequence $S$ is thus proportional to 
$ \prod_{\ell=1}^{k} e^{-\ln(\pi_\ell)\, E_i} = \prod_{\ell=1}^{k} \pi_i^{-E_i}. $

One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, \emph{i.e.} such that  $\mathbb{E}(E_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$.
The expected value of $E_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see App.~\ref{sec:weight_derivatives}). More generally, computing a suitable $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\cite{Denise2010,Bendkowski2017}. 
In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(exp(\beta),\dots,\exp(\beta))$ and, at each iteration, generates a sample $\mathcal{S}$ of sequences. The expected value of an energy $E_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}E_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$. 
$\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$. In practice, the constant $\gamma>1$ is chosen empirically to achieve effective optimization. 
While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior (choosing $\gamma=1.2$).

A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}. The rationale for a rejection approach follows from: 
i)\emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing Cupal \emph{et al}~\cite{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}). 
%Indeed, as shown in Figure~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
For base pair-based energy models and special compatibility graph
(paths, cycles\ldots) this property rigorously follows from analytic
combinatorics, see Bender-Richmond-Williamson~\cite{Bender1983} and
Drmota~\cite{Drmota1997}. In such cases, the expected number of
rejections before reaching the targeted energies remains constant when
$\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
$\varepsilon=0$. The \GCb-content of designs can also be controlled,
jointly and in a similar way, as done in
\Software{IncaRNAtion}~\cite{Reinharz2013}.

@@

For a multi-target design instance, individual reasonable target energies are estimated from averaging the energy over samples at relatively high weights $e^{\beta}$, controlling the \GCb-content at 50\%. For each instance of the \Software{Modena} benchmark~\cite{Taneda2015}, we estimated such target energies and subsequently generated $1\,000$ seed sequences at these energies.
As its rationale, this procedure is tailored to produce sequences with similar Turner energy that favor the stability of the target structures having moderate \GCb-content; all these properties are desirable objectives for RNA design.
All sampled sequences are evaluated under the multi-stable design objective function of \cite{Hammer2017}:
\vspace{-12pt}
% \begin{align}
%   f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
%    & +\ \xi \frac{2}{k(k-1)} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
% \end{align}
\begin{align}
  \label{eq:blueprintobjective}
    f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
   & +\ 0.5\frac{1}{\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
\end{align}
